Enzo: An Adaptive Mesh Refinement Code for Astrophysics

### 2013-03-15 introduction.tex

There are many diverse phenomena in astrophysics and cosmology which require a high spatial and temporal dynamical range. There are a number of numerical techniques which have been used to
model such problems.  The most common is a gridless, fully Lagrangian method call Smoothed Particle Hydrodynamcs \citep[SPH;][]{Lucy77, SPH} which has been very successful, particularly for those situations dominated by gravity.  Its development, however, is still at a relatively early stage when
compared to the effort put into Eulerian, grid-based hydrodynamic schemes \citep[e.g.,][]{Woodward84}.

Therefore, we would like to make use of this expertise; unfortunately, none of these schemes provide an easy way to adaptively and dynamically increase the resolution in small volumes, as required by, for example, gravitational instability.  We have instead, adapted an idea first suggested by \citet{Berger89} in the Computational Fluid Dynamics (CFD) community, which has become known as Structured Adaptive Mesh Refinement (SAMR).  The principle is to adaptively add (and modify) additional, finer meshes, or patches, over regions which require higher resolution.  Since each patch has a simple geometry (here, cartesian) the hyperbolic fluid-dynamical equations can be solved on that patch with a well-developed `off-the-shelf' numerical method.  In addition, it is relatively straightforward to add other advanced physics (e.g. MHD, radiation transport, etc.). This paper is devoted to describing how we have implemented this idea for astrophysics, including the addition of self-gravity, collisionless fluids, MHD, radiation transport and a range of other physical effects.

There have been many other numerical methods described in the astronomical literature which contain elements of this idea, or have a similar aim.  For example, an N-body solver \cite{Villumsen89} used non-adaptive meshing to increase the resolution in pre-selected regions.  This static approach, applied to hydrodynamics, has been used extensively \citep[e.g.,]{Ruffert94, Anninos94}.  Adding adaptivity is a more recent enhancement, and there are now a number of N-body codes which possess this feature, both with and without  hydrodynamics  \cite{Couchman91, Jessop94, Suisalu95, Splinter96, Gelato97, ART97, Truelove98, flash_method, MLAPM01,  Yahagi01, RAMSES, Quilis04, Ziegler05, Zhang06}.  Of these, perhaps the most comparable are FLASH \citep{flash_method}, which uses grid blocks of fixed size, and RAMSES \citep{RAMSES} and ART \citep{ART97}, both of which refine individual cells.  It is also possible to deform the grid to obtain high resolution \citep[e.g.,][]{Gnedin95, Xu97, Pen98}, and more recently

@@

The \enzo\ code has been widely used over the lasts decade in a wide variety of problems in topics ranging from star formation to turbulence to large-scale structure.  More than 100 peer-reviewed papers have been published based on results obtained with the code.

Finally, we note that the numerical simulation of astronomical phenomena now plays a key role, along with observations and analytic theory, in pushing forward our understanding of the cosmos.
%\citep[e.g.,][]{DecadalSurvey01}.  
But along with this role comes responsibility.  We believe that those developing simulation tools must fulfill two key obligations: the first is to make those tools available to the community as a whole, much in the way that astronomical data is now regularly made publicly available.  The second is document, test and refine those methods so that they can be critically evaluated and expanded by others.  This paper represents one part of our attempt to meet our obligations with respect to the \enzo\ code.

@@

%\red{
%\begin{enumerate}
%  \item Motivate why we're writing the paper now: (because there isn't one, lots of people use it for all sorts of things, given that it's publicly available we should really have a method paper and some general proof that the code works, and a list of what it does at this point in time)
%  \item Why have the \enzo\ code at all?  Many diverse phenomena in astrophysics and cosmology where you need a high spatial and temporal dynamical range.  Regular (i.e. monolithic) grids don't have the required range.  While SPH is the most commonly used tool, and has been quite successful, the lack of a grid structure complicates the usage of advanced hydrodynamical techniques and other advanced physics (e.g. full solution of the MHD equations, radiation transport beyond flux limited diffusion, etc.).  There has been 50(ish) years of advancement in the fields of grid-based solutions to these methods, and the use of adaptive mesh on structured grids allows us to take advantage of this for a wide range of astrophysical and cosmological applications.  
%  \item List the subjects that the code has been used for, citing appropriate papers.  Emphasize heavily landmark contributions (Greg's 1998 galaxy cluster paper and various Motl/Burns/Hallman cluster things, various lyman alpha forest stuff, Pop III papers, alexei's turbulence stuff, etc.).  This code has been used for approximately 10 years, and is publicly available.  List papers by non-collaborators that have already used the code.
%  \item Explain structure of the paper.  Emphasize that we pass through the enzo code in multiple levels of (increasing) detail: The ``physical equations'' section provides bare-bones ``what we're solving'' information, the ``overview of numerical methods'' section briefly explains (i.e. no pseudocode or finite difference stencils) how the various important subsystems of the code work, and the appendices of the paper have the real nitty-gritty details that most people don't want to know, but a select few will find highly useful.
%\end{enumerate}
%} %end red

>>>>

Many phenomena in astrophysics and cosmology require a high spatial and temporal dynamical range and there are a number of numerical techniques which have been used to model such problems.  At the current time, the most common is a gridless, fully Lagrangian method call Smoothed Particle Hydrodynamcs \citep[SPH;][]{Lucy77, SPH} which has been very successful, particularly for those situations dominated by gravity.  Its development, however, is still at a relatively early stage when compared to the effort put into Eulerian, grid-based hydrodynamic schemes \citep[e.g.,][]{, laney-1998, toro-1997, Woodward84}.

Therefore, we would like to make use of this expertise that has been invested in Eulerian solvers; unfortunately, as originally developed, none of these schemes provide an easy way to adaptively and dynamically increase the resolution in small volumes, as required by, for example, gravitational instability.  We have instead, adopted an idea first suggested by \citet{Berger89} in the Computational Fluid Dynamics (CFD) community, which has become known as Structured Adaptive Mesh Refinement (SAMR).  The principle is to adaptively add (and modify) additional, finer meshes, or patches, over regions which require higher resolution.  Since each patch has a simple geometry (here, Cartesian) the hyperbolic fluid-dynamical equations can be solved on that patch with a well-developed `off-the-shelf' numerical method.  In addition, it is relatively straightforward to add other advanced physics (e.g. MHD, radiation transport, etc.). This paper is devoted to describing how we have implemented this idea for astrophysics, including the addition of comoving coordinates, self-gravity, radiative cooling, chemistry, heat conduction, collisionless fluids, MHD, radiation transport, star formation and a range of other physical effects.

There have been many other numerical methods described in the astronomical literature which contain elements of this idea, or have a similar aim.  For example, an N-body solver \cite{Villumsen89} used non-adaptive meshing to increase the resolution in pre-selected regions.  This static approach, applied to hydrodynamics, has been used extensively \citep[e.g.,]{Ruffert94, Anninos94}.  Adding adaptivity is a more recent enhancement, and there are now a number of N-body codes which possess this feature, both with and without  hydrodynamics  \citep{Couchman91, Jessop94, Suisalu95, Splinter96, Gelato97, ART97, Truelove98, flash_method, MLAPM01,  Yahagi01, RAMSES, Quilis04, Ziegler05, Zhang06, Astrobear09, Pluto-amr, GAMER, Nyx}.  Of these, perhaps the most comparable and widely used are FLASH \citep{flash_method}, which uses grid blocks of fixed size, and RAMSES \citep{RAMSES} and ART \citep{ART97}, both of which refine individual cells.   It is also possible to deform the grid to obtain high resolution \citep[e.g.,][]{Gnedin95, Xu97, Pen98}, and more recently a few grid codes have adopted an unstructured approach based on a moving Voronoi mesh \citep{Arepo10, Tess11}.

@@

The \enzo\ code has been widely used over the last two decades in a wide variety of problems, and more than 100 peer-reviewed papers have been published based on results obtained with the code.  The code has previously been used to study a wide range of astrophysical systems, including X-ray clusters \citep{Loken02}, the ISM \citep{Slyz05}, the intergalactic medium \citep{Fang01}, cooling flows \citep{Li12}, turbulence \citep{Kritsuk04}, and the formation of the first stars \citep{ABN02}.

Finally, we note that the numerical simulation of astronomical phenomena now plays a key role, along with observations and analytic theory, in pushing forward our understanding of the cosmos \citep[e.g.,][]{DecadalSurvey01, DecadalSurvey10}.   But along with this role comes responsibility.  We believe that those developing simulation tools must fulfill two key obligations: the first is to make those tools available to the community as a whole, much in the way that astronomical data is now regularly made publicly available.  The second is document, test and refine those methods so that they can be critically evaluated and expanded by others.  This paper represents one part of our attempt to meet our obligations with respect to the \enzo\ code.

@@

### 2013-03-15 numerical-gravity.tex

\subsection{Gravity: Solving the Poisson Equations}
\label{sec.gravity}

There are multiple methods to compute the gravitational potential (which is an elliptic equation in the Newtonian limit) in a structured AMR framework.  One way would be to model the dark matter (or other collisionless particle-like objects, such as stars) as a second fluid in addition to the baryonic fluid and solve the collisionless Boltzmann equation, which follows the evolution of the fluid density in six-dimensional phase space.  However, this is computationally prohibitive owing to the large dimensionality of the problem, making this approach unappealing for the cosmological AMR code.

The dark matter particles are distributed onto the grids using the cloud-in-cell (CIC) interpolation technique to form a spatially discretized density field (analogous to the baryon densities used to calculate the equations of hydrodynamics).  After sampling dark matter density onto the grid and adding baryon density if it exists (to get the total matter density in each cell), the gravitational potential is calculated on the periodic root grid using a fast Fourier transform.  For periodic boundary conditions, we can use either a simple Greens function kernel of $-k^{-2}$, or the finite-difference equivalent \citep{Hockney88}:
\begin{equation}
G(\vec{k}) = - \frac{\Delta x}{2 \left( sin(k_x \Delta x/2)^2 + sin(k_y \Delta y/2)^2 + sin(k_z \Delta z/2)^2 \right) }
\end{equation}
where $k^2 = k_x^2 + k_y^2 + k_z^2$ is the wavenumber in Fourier space and the potential is calculated in k-space as usual with $\tilde{\phi}(k) = G(k) \tilde{\rho}(k)$.

For isolated boundary conditions, we use James method (REF).  In this case, the Greens function is generated in real-space so as to have the correct zero-padding properties and then transformed into the Fourier domain.

>>>>

\subsection{Gravity}
\label{sec.gravity}

Solving for the accelerations for the cells and particles in a grid due to self-gravity involves three steps: first, computing the total gravitating mass; second, solving for the potential field with the appropriate boundary conditions, and finally, differencing the potential to get the acceleration and, if necessary, interpolating the potential back to the particles. These steps are described in detail below.

First, the dark matter particles are distributed onto the grids using the cloud-in-cell (CIC) interpolation technique \citet{Hockney88} to form a spatially discretized density field $\rho_{\rm DM}$.  When doing the CIC interpolation, particle positions are (temporarily) advanced by $0.5 \Delta t v^n$ so that we generate an estimate of the time-centered density field.  Particles on sub-grids are also added in this way.  In addition, since the gravitating field for a grid is defined beyond the grid edges (see below), dark matter particles from sibling grids and sub grids which lie within the entire gravitating field are used.

Next, we add the baryonic grid densities in a similar fashion.  In particular, we treat baryonic cells as virtual CIC particles which are are placed at the grid center but are advanced by $0.5 \Delta t v^n$ in order to approximately time-center the gravitating mass field.  Cells which are covered by further refined grids are treated in a similar way (i.e. we use the sub grid cells as virtual CIC particles).  This procedure results in a total gravitating mass field $\rho_{\rm total}^{n+1/2}$.

To compute the potential field from this gravitating mass field on the root grid, we use a a fast Fourier transform. For periodic boundary conditions, we can use either a simple Greens function kernel of $-k^{-2}$, or the finite-difference equivalent \citep{Hockney88}:
\begin{equation}
G(\vec{k}) = - \frac{\Delta x}{2 \left( sin(k_x \Delta x/2)^2 + sin(k_y \Delta y/2)^2 + sin(k_z \Delta z/2)^2 \right) }
\end{equation}
where $k^2 = k_x^2 + k_y^2 + k_z^2$ is the wavenumber in Fourier space and the potential is calculated in k-space as usual with $\tilde{\phi}(k) = G(k) \tilde{\rho}(k)$.

For isolated boundary conditions, we use James method (REF).  In this case, the Greens function is generated in real-space so as to have the correct zero-padding properties and then transformed into the Fourier domain.  In either case, the potential is then transformed back into the real domain to get potential values at the cell centers.  These are differenced with a two-point centered difference to get accelerations at the cell centers (except if we are using the staggered Zeus-like solver, in which case the accelerations are computed at the cell faces to match the velocities).
